---
title: "SRKW-QERM514"
author: "Mollie Ball"
date: "2026-05-08"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## Sets up the Environment and Libraries

# Probably optional: packages to run Phyloseq code 
##(for wrangling DADA2 outputs)

# if(!requireNamespace("BiocManager")){
#   install.packages("BiocManager")
# }
# BiocManager::install("phyloseq")

# library(phyloseq); packageVersion("phyloseq")
# library(Biostrings); packageVersion("Biostrings")

# required packages
library(ggplot2); packageVersion("ggplot2")
library(tidyverse)
library(dplyr)
library(tibble)
library(patchwork)
library(MASS)
library(lme4)
library(glmmTMB)
library(MuMIn)
library(DHARMa)
library(lubridate)
```

## Fine Scale Diet Variability in Southern resident killer whales (Orcinus orca ater)
	Southern resident killer whales (Orcinus orca ater; SRKW), with just 74 members of the population, have been listed as Endangered Species under the U.S. Endangered Species Act (ESA) since 2005 (Krahn et al., 2004). They occupy the inland and offshore waters of the Pacific Northwest, with ranges spanning from California to Southern Alaska; critical habitat for SRKWs includes Puget Sound, found in Washington state (Krahn et al., 2004).
	The main threats to SRKW are a lack of prey, vessel disturbance, contaminants (especially persistent organic pollutants (POPs)), and inbreeding (Switzer et al., 2026; Van Cise et al., 2024). Further, nutritional stress can work synergistically to amplify the effects of each stressor (Switzer et al., 2026). SRKW diets are culturally learned and believed to comprise almost entirely (>80%) of Chinook salmon (Oncorhynchus tshawytscha) (Van Cise et al., 2024). Many Chinook salmon populations in the Puget Sound region are also listed as threatened, endangered, or critical under the Endangered Species Act (Ford et al., 2011). 
	For this project, I therefore aim to use data generated by genetic metabarcoding of the 16S region of mitochondrial DNA SRKW fecal samples collected from 2007 to 2025 (n=nearly 500[MB1.1]) in Washington state and surrounding waters to describe any spatial, temporal, and/or demographic (e.g. by sex, pod, age class, etc.) changes therein. [MB2.1]Library preparation was conducted by the San Diego Zoo and Wildlife Alliance (Escondido, CA). [MB3.1]
Research Question: 
Hypotheses and Expectations: hypothesis/hypotheses you’re addressing; predictions from hypotheses
Hypothesis 1: There will be variability in SRKW diet across pods (J, K, and L)
Hypothesis 2: There will be variability in SRKW diet across year
Hypothesis 3: There will be variability in SRKW diet across sex
Hypothesis 4: There will be variability in SRKW diet across season

Preliminary analysis (175 samples out of ~500 total samples[MB4.1]) shows that the diet composition of SRKW is relatively consistent with known diet composition, with over 80% of sequence reads assigning to Onchorhynchus tshawytascha (Figure 4). 

## Background about these data
Data generated by genetic metabarcoding of the 16S region of mitochondrial DNA SRKW fecal samples collected from 2007 to 2025 in Washington state and surrounding waters. 


```{r Loads in Data}
getwd()
# Loads in output from DADA2 + filtered seqtab.nochim and samdf from rownames-match.r
load("replicates-contaminated.RData")

# Loads in species relative abundance x sample table
df <- read.csv("./Deliverables/ALL/SRKW_relative_speciesxsamples.csv")
```


WHY DO THEY NEED TO BE RELATIVE ABUNDANCE AND NOT ABSOLUTE ABUNDANCE?
```{r Formats the table for ease of use}
#?select()

# Messes with row and column names in samdf_filt for ease
samdf_filt <- samdf_filt %>%
  rename(Sample_name2 = Sample_name)%>%
  rownames_to_column("Sample_name")

# Selects the column for relative abundance of Chinook per sample
# renames column X to match samdf_filt
head(df)

df_chin <- df %>%
  dplyr::select(X, Oncorhynchus.tshawytscha)%>%
  dplyr::rename(Sample_name = X) %>%
  dplyr::rename(chin = Oncorhynchus.tshawytscha)

# Adds metadata from samdf_filt to df_chin by Sample_name
?left_join()
df_chin <- left_join(df_chin, samdf_filt, by = "Sample_name")

# Cleans Pod column
df_chin$Pod[df_chin$Pod == "unknown"] <- "UNK"
df_chin$Pod[df_chin$Pod == is.na(df_chin$Pod)] <- "UNK"

```


$$ DESCRIBE THE DATA HERE $$
' chin = the relative abundance of Chinook in the diet of each sample (functionally, the proportion of Chinook salmon)


I WANT TO LOOK AT THE PERCENT OF CHINOOK IN THE DIET ACROSS DIFFERENT PREDICTOR VARIABLES
SEASON
AGE CLASS
SEX
POD
YEAR

## Including Plots

You can also embed plots, for example:

```{r Exploratory plots}

# Plots the relative abundance of Chinook per sample
p1 <- ggplot(df_chin, aes(ID, chin, color = ID)) + 
  geom_point()+
  ylab("Relative Abundance of Chinook")+
  theme_minimal()+
  theme(legend.position = "none")
  
p1

# Plots the relative abundance of Chinook per year
p2 <- ggplot(df_chin, aes(Year, chin, color = ID)) + 
  geom_point() + 
  ylab("Relative Abundance of Chinook")+
  theme_minimal()+
  theme(legend.position = "none")
  
p2

# Plots the relative abundance of Chinook per season
p3 <- ggplot(df_chin, aes(julian_day, chin, color = ID)) + 
  geom_point()+ 
  # ylab("Relative Abundance of Chinook (%)")+
  theme_minimal() +
  theme(legend.position = "none")
  
p3

# Plots the relative abundance of Chinook per pod
p4 <- ggplot(df_chin, aes(Pod, chin, color = ID)) + 
  geom_point()+ 
  # ylab("Relative Abundance of Chinook (%)")+
  theme_minimal()+
  theme(legend.position = "none")
  
p4

(p1 + p2) / (p3 + p4)


## I DONT HAVE THIS DATA CLEANED UP YET
# # Plots the relative abundance of Chinook per sex
# p5 <- ggplot(df_chin, aes(sex, chin)) + 
#   geom_point()+ 
#   theme_minimal()
#   
# p5

```

According to these plots, I see some possible differences across parameters. 
For individual, I see that there are many whales for which we have more than one sample. I also see a wide spread of percent of chinook found in the diet (0%-100% chinook). Because there is individual variation here, I want to make sure to capture that in the model. Because each sample from the same individual will not be independent (samples from one individual can be thought of as a "grouping"), I plan to include this as a random effect. 

I don't see a clear pattern for relative abundance of chinook in the diet by year. 

I also don't see any pattern for relative abundance of chinook by season, expect that it looks like sampling effort was lower in the Spring compared to the Summer, Fall, and Winter. I also see that individuals from different pods may be more prevalent in our sampling areas in certain seasons. For example, individuals from J pod seem to make up most of the available samples from the summer months, and samples from L pod may be more available in the Autumn. Interestingly, there may also be a higher percent of Chinook in the diet for L pod in Autumn vs J pod, but it is hard to make any real statements at this time. 


## Now, to explore what kind of model to run, I will further explore the data and run some assumptions tests. 


```{r tests for normality}

# Visual and statistical tests for normality

# Histogram of the relative abundance of chinook
?hist()
hist(df_chin$chin) #this data definitely does not appear normal


# Shapiro Wilkes test for normality

# makes chin in df_chin an numeric vector
x <- df_chin$chin
range(x)

# Runs Shapiro-Wilk normality test
## Null hypothesis: the data are normal
## Alt hypotheeiss: the data are not normal
## alpha = 0.5

shapiro.test(x) #p = 2.2e-16

```
Upon visual inspection, the data does not look normally distributed. Using a Shapiro-Wilk test, this is corroborated. With a p value << 0.05, we reject the null and conclude that the data are not normally distributed. 

This rules out using 

Simple model: does the relative abundance of Chinook in the diet of SRKW vary by pod?

Because of the distribution of this data, and because it is proportional data, I have chosen to use a GLMM with a beta regression and the logit link function. Here are the three fundamental components: 

distribution of the data:

a_i and beta_i is a transformation that gets you to the expectation of P= ?
$$
P_i \sim Beta(\alpha_i, \beta_i)
$$
link function: maybe don't need the logit contraint?
what is the expectation for a beta sistribution
$$
logit(\alpha_i)=log(\alpha_i/1−\alpha_i)= \eta_i
$$

linear predictor: 
alpha_i = pod

eta = expectation of the beta
$$
\eta_i = \beta_0 +\beta_1\alpha_i

$$
COULD JUST DO A LINEAR REGRESSION WITH A LOGIT TRANSFORM ON THE P (BETA

lmer(y ~ x)
where y is logit_p

```{r tests for the simple model for chin~pod (no random effect)}


## even though my range was confirmed as 0 - 1, glmmTMB hates it, stackoverflow said to try this 'nudge' method.
eps <- 0.001
df_chin$chin_beta <- pmin(pmax(df_chin$chin, eps), 1 - eps)

pod_mod <- glmmTMB(chin_beta ~ Pod,
                   data = df_chin,
                   family = beta_family(link = "logit"))
summary(pod_mod)

## USE DHARMa FOR DIAGNOSTIC PLOTS

```
``` {r pod res}
# Simulate residuals
res_pod <- simulateResiduals(
  fittedModel = pod_mod,
  n = 1000,          # more sims = smoother diagnostics
  plot = FALSE       # i want to plot in a separate line
)

# Basic DHARMa diagnostic plots
plot(res_pod)        # QQ + residual vs predicted + histogram, etc.

# Uniformity test (are residuals ~ Uniform(0,1)?)
testUniformity(res_pod)

# Over/underdispersion test
testDispersion(res_pod)

# Outlier test
testOutliers(res_pod)

# residuals vs fitted values

plotResiduals(res_pod, predict(pod_mod, type = "response"))
```

Nothing but the intercept is significant here, showing that there is no difference in the relative abundance of salmon across pod (not including a random effect)

with the random effect: 

distribution of the data:
$$
X_i \sim Beta(\alpha_i, \beta_i)
$$
link function:
$$
logit(\alpha_i)=log(\alpha_i/1−\alpha_i)= \eta_i
$$

linear predictor: 
alpha_i = pod
delta_i = RE for ID
$$
\eta_i = \beta_0 + \beta_1\alpha_i + \delta_i
$$


```{r tests for the simple model for chin~pod with random effect}
pod_wRE_mod <- glmmTMB(chin_beta ~ Pod + (1 | ID),
                   data = df_chin,
                   family = beta_family(link = "logit"))
summary(pod_wRE_mod)

```
Still, nothing aside from the intercept is significant. 


``` {r pod wRE res}

?glmmTMB()

# Simulate residuals
res_pod_wRE <- simulateResiduals(
  fittedModel = pod_wRE_mod,
  n = 1000,          # more sims = smoother diagnostics
  plot = FALSE       # we'll call plot() explicitly
)

# Basic DHARMa diagnostic plots
plot(res_pod_wRE)

# Uniformity test (are residuals ~ Uniform(0,1)?)
testUniformity(res_pod_wRE)

# Over/underdispersion test
testDispersion(res_pod_wRE)

# Outlier test
testOutliers(res_pod_wRE)

# residuals vs fitted values

plotResiduals(res_pod_wRE, predict(pod_wRE_mod, type = "response"))
```


Let's try year:

distribution of the data:
$$
X_i \sim Beta(\alpha_i, \beta_i)
$$
link function:
$$
logit(\alpha_i)=log(\alpha_i/1−\alpha_i)= \eta_i
$$

linear predictor: 
alpha_i = year
$$
\eta_i = \beta_0 + \beta_1\alpha_i
$$


```{r tests for the simple model for chin~year (no random effect)}


year_mod <- glmmTMB(chin_beta ~ Year,
                   data = df_chin,
                   family = beta_family(link = "logit"))
summary(year_mod)

```
Year is not significant. 

with the random effect: 

distribution of the data:
$$
X_i \sim Beta(\alpha_i, \beta_i)
$$
link function:
$$
logit(\alpha_i)=log(\alpha_i/1−\alpha_i)= \eta_i
$$

linear predictor: 
alpha_i = year
delta_i = RE for ID
$$
\eta_i = \beta_0 + \beta_1\alpha_i + \delta_i
$$

```{r tests for the simple model for chin~year with random effect}
year_wRE_mod <- glmmTMB(chin_beta ~ Year + (1 | ID),
                   data = df_chin,
                   family = beta_family(link = "logit"))
summary(year_wRE_mod)


```

Still nothing. 

Trying season:
$$
X_i \sim Beta(\alpha_i, \beta_i)
$$
link function:
$$
logit(\alpha_i)=log(\alpha_i/1−\alpha_i)= \eta_i
$$

linear predictor: 
alpha_i = julian day
$$
\eta_i = \beta_0 + \beta_1\alpha_i
$$

```{r tests for the simple model for chin~season (no random effect)}


jul_mod <- glmmTMB(chin_beta ~ julian_day,
                   data = df_chin,
                   family = beta_family(link = "logit"))
summary(jul_mod)

```

We are cooking with something now! Some of the months are significant. 



$$
X_i \sim Beta(\alpha_i, \beta_i)
$$
link function:
$$
logit(\alpha_i)=log(\alpha_i/1−\alpha_i)= \eta_i
$$

linear predictor: 
alpha_i = julian day
delta_i = RE for ID
$$
\eta_i = \beta_0 + \beta_1\alpha_i + \delta_i
$$

adding a random effect:
```{r tests for the simple model for chin~season with random effect}


jul_wRE_mod <- glmmTMB(chin_beta ~ julian_day + (1 | ID),
                   data = df_chin,
                   family = beta_family(link = "logit"))
summary(jul_wRE_mod)

```


I now want to see if we can make this better: 

```{r tests for the simple model for chin~julian day with an interaction term?}

yearjul_wRE_mod <- glmmTMB(chin_beta ~ julian_day + Year + (1 | ID),
                   data = df_chin,
                   family = beta_family(link = "logit"))
summary(yearjul_wRE_mod)

```


```{r tests for the simple model for chin~season with random effect}

podmonth_mod <- glmmTMB(chin_beta ~ Pod + julian_day,
                   data = df_chin,
                   family = beta_family(link = "logit"))
summary(podmonth_wRE_mod)

```


```{r tests for the simple model for chin~season with random effect}

podmonth_wRE_mod <- glmmTMB(chin_beta ~ Pod + julian_day + (1 | ID),
                   data = df_chin,
                   family = beta_family(link = "logit"))
summary(podmonth_wRE_mod)

```
SAY SOMETHING ABOUT THAT PROBABLY



``` {r for getting a delta AIC table}
## Collect models in a named list
mod_list <- list(
  pod      = pod_mod,
  pod_RE   = pod_wRE_mod,
  year     = year_mod,
  year_RE  = year_wRE_mod,
  jul    = jul_mod,
  jul_RE = jul_wRE_mod, 
  podmonth = podmonth_mod,
  podmonth_RE = podmonth_wRE_mod,
  podmonth_2RE = podmonth_w2RE_mod
  
)

## Compute AICc for each
mod_res <- data.frame(
  model = names(mod_list),
  AICc  = sapply(mod_list, MuMIn::AICc)
)

## Order by AICc
mod_res <- mod_res[order(mod_res$AICc), ]

## ΔAICc
mod_res$delta_AICc <- mod_res$AICc - min(mod_res$AICc)

## Akaike weights based on AICc
mod_res$weight <- exp(-0.5 * mod_res$delta_AICc)
mod_res$weight <- mod_res$weight / sum(mod_res$weight)

## Round for easier viewing
mod_res$AICc        <- round(mod_res$AICc, 2)
mod_res$delta_AICc  <- round(mod_res$delta_AICc, 2)
mod_res$weight      <- round(mod_res$weight, 2)

mod_res
```

The best model here is actually the model that just includes julian day (and not a random effect for individual). However, I don't think that it makes sense to move forward with a model without adding a random effect for individual, since there are multiple samples from the same individuals across time. I wonder also if a random effect for year would be appropriate, because the availability of salmon for them to eat would definitely be related (because )

YOLO let's try it:

```{r tests for the simple model for chin~season with random effect}

podmonth_w2RE_mod <- glmmTMB(chin_beta ~ Pod + julian_day + (1 | ID) + (1 | Year),
                   data = df_chin,
                   family = beta_family(link = "logit"))
summary(podmonth_w2RE_mod)

```


Let's run some diagnostics, though, for funzies.


Checks of random‑effects variance estimates, overdispersion, and leverage/influence.
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
