---
title: "SRKW-QERM514"
author: "Mollie Ball"
date: "2026-05-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## Sets up the Environment and Libraries

# Probably optional: packages to run Phyloseq code 
##(for wrangling DADA2 outputs)

# if(!requireNamespace("BiocManager")){
#   install.packages("BiocManager")
# }
# BiocManager::install("phyloseq")

# library(phyloseq); packageVersion("phyloseq")
# library(Biostrings); packageVersion("Biostrings")

# required packages
library(ggplot2); packageVersion("ggplot2")
library(tidyverse)
library(dplyr)
library(tibble)
library(patchwork)
```

## Fine Scale Diet Variability in Southern resident killer whales (Orcinus orca ater)
	Southern resident killer whales (Orcinus orca ater; SRKW), with just 74 members of the population, have been listed as Endangered Species under the U.S. Endangered Species Act (ESA) since 2005 (Krahn et al., 2004). They occupy the inland and offshore waters of the Pacific Northwest, with ranges spanning from California to Southern Alaska; critical habitat for SRKWs includes Puget Sound, found in Washington state (Krahn et al., 2004).
	The main threats to SRKW are a lack of prey, vessel disturbance, contaminants (especially persistent organic pollutants (POPs)), and inbreeding (Switzer et al., 2026; Van Cise et al., 2024). Further, nutritional stress can work synergistically to amplify the effects of each stressor (Switzer et al., 2026). SRKW diets are culturally learned and believed to comprise almost entirely (>80%) of Chinook salmon (Oncorhynchus tshawytscha) (Van Cise et al., 2024). Many Chinook salmon populations in the Puget Sound region are also listed as threatened, endangered, or critical under the Endangered Species Act (Ford et al., 2011). 
	For this project, I therefore aim to use data generated by genetic metabarcoding of the 16S region of mitochondrial DNA SRKW fecal samples collected from 2007 to 2025 (n=nearly 500[MB1.1]) in Washington state and surrounding waters to describe any spatial, temporal, and/or demographic (e.g. by sex, pod, age class, etc.) changes therein. [MB2.1]Library preparation was conducted by the San Diego Zoo and Wildlife Alliance (Escondido, CA). [MB3.1]
Research Question: 
Hypotheses and Expectations: hypothesis/hypotheses you’re addressing; predictions from hypotheses
Hypothesis 1: There will be variability in SRKW diet across pods (J, K, and L)
Hypothesis 2: There will be variability in SRKW diet across seasons 
Hypothesis 3: There will be variability in SRKW diet across sex
Hypothesis 4: There will be variability in SRKW diet across time

Preliminary analysis (175 samples out of ~500 total samples[MB4.1]) shows that the diet composition of SRKW is relatively consistent with known diet composition, with over 80% of sequence reads assigning to Onchorhynchus tshawytascha (Figure 4). 

## Background about these data
Data generated by genetic metabarcoding of the 16S region of mitochondrial DNA SRKW fecal samples collected from 2007 to 2025 in Washington state and surrounding waters. 


```{r Loads in Data}

# Loads in output from DADA2 + filtered seqtab.nochim and samdf from rownames-match.r
load("rownames-match.RData")

# Loads in species relative abundance x sample table
df <- read.csv("./Deliverables/ALL/SRKW_relative_speciesxsamples.csv")
```


WHY DO THEY NEED TO BE RELATIVE ABUNDANCE AND NOT ABSOLUTE ABUNDANCE?
```{r Formats the table for ease of use}
#?select()

# Messes with row and column names in samdf_filt for ease
samdf_filt <- samdf_filt %>%
  rename(Sample_name2 = Sample_name)%>%
  rownames_to_column("Sample_name")

# Selects the column for relative abundance of Chinook per sample
# renames column X to match samdf_filt
df_chin <- df %>%
  select(X, Oncorhynchus.tshawytscha)%>%
  rename(Sample_name = X) %>%
  rename(chin = Oncorhynchus.tshawytscha)

# Adds metadata from samdf_filt to df_chin by Sample_name
?left_join()
df_chin <- left_join(df_chin, samdf_filt, by = "Sample_name")

# Add a column for season
df_chin <- df_chin %>%
  mutate(season = case_when(
    Month %in% c("Dec", "Jan", "Feb") ~ "Winter",
    Month %in% c("Mar", "Apr", "May") ~ "Spring",
    Month %in% c("Jun", "Jul", "Aug") ~ "Summer",
    Month %in% c("Sep", "Oct", "Nov") ~ "Autumn"
  ))

```

I WANT TO LOOK AT THE PERCENT OF CHINOOK IN THE DIET ACROSS DIFFERENT PREDICTOR VARIABLES
SEASON
AGE CLASS
SEX
POD
YEAR

## Including Plots

You can also embed plots, for example:

```{r Exploratory plots}

# Plots the relative abundance of Chinook per sample
p1 <- ggplot(df_chin, aes(ID, chin)) + 
  geom_point()
  
p1

# Plots the relative abundance of Chinook per year
p2 <- ggplot(df_chin, aes(Year, chin, fill = ID)) + 
  geom_point() + 
  theme_minimal()+
  theme(legend.position = "none")
  
p2

# Plots the relative abundance of Chinook per season
p3 <- ggplot(df_chin, aes(season, chin)) + 
  geom_point()+ 
  theme_minimal() 
  
p3

# Plots the relative abundance of Chinook per pod
p4 <- ggplot(df_chin, aes(pod, chin)) + 
  geom_point()+ 
  theme_minimal()
  
p4

(p1 + p2) / (p3 + p4)


## I DONT HAVE THIS DATA YET
# # Plots the relative abundance of Chinook per sex
# p5 <- ggplot(df_chin, aes(sex, chin)) + 
#   geom_point()+ 
#   theme_minimal()
#   
# p5

```

Reminder: the models I am interested in
Hypothesis 1: There will be variability in SRKW diet across pods (J, K, and L)
Hypothesis 2: There will be variability in SRKW diet across seasons 
Hypothesis 3: There will be variability in SRKW diet across sex
Hypothesis 4: There will be variability in SRKW diet across time

```{r m1 - percent chinook in diet vs pod}

```
```{r m4 - percent chinook in diet vs season}

```
```{r m1 - percent chinook in diet vs sex}

```
```{r m1 - percent chinook in diet vs year}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
